A Transdiciplinary approach to developing a 3D spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities  around the coast of South Africa

### 2019-10-16 literature_review.Rmd
some grammer spotted and corrected

The morphological adaptation that macroalgae display are therefore driven by site conditions, therefore individuals must be morphologically flexible to persist in stochastic environments. This may be achieved through different strategies and are species-specific, which can be directly attributed the high diversity in morphological characters of algal communities. For example some algae have fronds and others are articulated coralline, and therefore these species would need to adapt their morphology differently in order to persist. In general, flat strap-like blades are common in areas that are exposed to high wave energy, while at protected sites blade morphology is wide and undulated.

@@

Waves are complex and various approaches, techniques and devices have been developed in order to measure waves effectively. One of the ways that has been used extensively used in the past but less so today, are visual estimates which can be used to characterise *significant swell height* (Hs) and *significant swell period* (Tp). Visual estimates use 15 - 20 of the most well defined, higher waves of a number of wave groups to characterise Hs and Tp. Although these parameters are useful, they do not accurately reflect the waves that are occurring in nature. Ocean waves are a combination of wind sea (short, irregular, locally generated waves) and swell (long, smooth waves, generated by distant storms) and so more parameters are needed to separate Hs and Tp driven by wind sea and swell, i.e. separate Hs and Tp parameters for wind sea and swell. However, even with separate parameters for the different types of waves would still not be enough to effectively characterise waves in the ocean. In order to characterise the detail and complexity of ocean waves a different approach must be used, know as the *spectral* approach. This approach is based on the idea that the sea surface can be characterised as the summation of a large number of harmonic wave components, known as the *random-phase/amplitude* model. The random-phase amplitude model is a summation of wave components on a discrete scale, however a continous scale is more relevant to characterising and measuring of waves in the ocean. The discrete spectrums are converted to continoius spectrums through various methods. Therefore, the random-phase amplitude model is used to obtain a *continious variance density spectrum* which is a statistical relevant measure for scientists and engineers.

The wave spectrum is based on several spectra, namely the amplitude spectrum, variance spectrum, variance density spectrum (discontinuous) and the variance density spectrum (continuous). A wave spectrum can be used to describe the sea surface as a stochastic process, however two conditions (assumptions) must first be met 1) the observation must be stationary and 2) the surface elevations must be Gaussian distributed. As mentioned previously, the random-phase/amplitude model approach characterises waves as harmonic and the time-record can be reproduced by the sum of a large number of harmonic waves using a *Fourier analysis*. A Fourier analysis allows values of the amplitude and phase for each frequency to be characterised, which in turn provides the amplitude and phase spectrum for a single time-record. The amplitude spectrum is can be calculated by averaging many time-records which determines the *average amplitude spectrum*. This approach allows the entire wave record to be characterised but is not statistically relevant and therefore is not applicable to hydrodynamic modelling. In order for to attain a statistically relevant value, the variance of each wave component must be calculated. The variance is regarded as statistically relevant as the sum of the variances (i.e. random surface elevation) of the wave components is equal to the variance of the sum of the wave components. In addition, LWT dictates that the energy of waves is proportional to the variance and is therefore an indirect measure of wave energy. This indirect measurement can also be used to determine other important wave components such as wave-induced particle velocity and pressure variations. Important to note that the measurement of variance are discrete and is therefore limited in characterising waves. A continuous scale is more applicable for hydrodynamic modelling purposes and is also closer to representing ocean waves and can be calculated by allowing the frequency interval () to approach zero.

@@

Wave exposure may be modeled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorporates integrative measures of hydrodynamic conditions at a particular site. Cartographic models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurements for larger areas, and has been suggested as a method for predicting macroalgal community structure (???). An example of such a model is the “BioEx model” which was developed by Baardseth and others (1970) to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiversity (???). Lindegarth and Gamfeldt (2005) criticised this approach, arguing that the choice of wave exposure method can influence ecological inference. The authors also highlighted the need for objective, reproducible and quantitative studies comparing exposure indices (Lindegarth and Gamfeldt 2005). Other authors, such as Hill et al. (2010), have argued that these simple measures can be improved upon by including bathymetry data which allows the incorporation of diffraction into the calculation. Diffraction is topographically induced variations in wave direction. A model incorporating this complexity was developed by Isæus (2004), and is known as the “simplified wave model” (SWM). The model uses measurements of wind strength, fetch and empirically derived algorithms to mimic diffraction.

Advances in numerical modelling have been founded on physical wave theory on how a wave "behaves". This approach is based on a theoretical perspective rather than the need to answer ecological questions. Besides diffraction, numerical models incorporate more complexity by including wind forcing, wave-to-wave interactions and loss of energy due to friction and wave breaking. Numerical models have a variety of applications and are often incorporated within hydrodynamic general circulation models and are used operationally for forecasting the sea state (Group 1988; Booij, Ris, and Holthuijsen 1999; Smith, Sherlock, and Resio 2001). The downside of advanced numerical models is that are computationally intensive which creates limitations for large scale simulations. Therefore, their application along long stretches of variable coastline, inshore environments and ocean-wide simulations are limited due to the poor spatial coverage. However, numerical models can be designed for local or site specific coverage, provided the correct data is available. **...More to be added...**

>>>>

The morphological adaptation that macroalgae display are driven by site conditions, therefore individuals must be morphologically flexible to persist in stochastic environments. This may be achieved through different strategies and are species-specific, which can be directly attributed the high diversity in morphological characters of algal communities. For example some algae have fronds and others are articulated coralline, and therefore these species would need to adapt their morphology differently in order to persist. In general, flat strap-like blades are common in areas that are exposed to high wave energy, while at protected sites blade morphology is wide and undulated.

# The mechanisms of morphological adaptation
## Waves and macroalgae characteristics

In the wave swept nearshore one would expect organisms to reflect the harsh hyrdrodyanmic environment by being streamlined, small and amoured. This is certianly the case for an array of fauna which often comprise of hard, rigid bodies that are held firmly in place to the substratum such as limpets and isopods.

@@

Waves are complex and various approaches, techniques and devices have been developed in order to measure waves effectively. One of the ways that has been used extensively used in the past but less so today, are visual estimates which can be used to characterise *significant swell height* (Hs) and *significant swell period* (Tp). Visual estimates use 15 - 20 of the most well defined, higher waves of a number of wave groups to characterise Hs and Tp. Although these parameters are useful, they do not accurately reflect the waves that are occurring in nature. Ocean waves are a combination of wind sea (short, irregular, locally generated waves) and swell (long, smooth waves, generated by distant storms) and so more parameters are needed to separate Hs and Tp driven by wind sea and swell, i.e. separate Hs and Tp parameters for wind sea and swell. However, even with separate parameters for the different types of waves would still not be enough to effectively characterise waves in the ocean. In order to characterise the detail and complexity of ocean waves a different approach must be used, know as the *spectral* approach. This approach is based on the idea that the sea surface can be characterised as the summation of a large number of harmonic wave components, calculated by the *random-phase/amplitude* model. The random-phase amplitude model is a summation of wave components on a discrete scale, however a continous scale is more relevant to characterising and measuring of waves in the ocean. The discrete spectrums are converted to continoius spectrums through various arithmetic methods. Therefore, the random-phase amplitude model is used to obtain a *continious variance density spectrum* which is a statistical relevant measure for scientists and engineers.

The wave spectrum is based on several spectra, namely the amplitude spectrum, variance spectrum, variance density spectrum (discontinuous) and the variance density spectrum (continuous). A wave spectrum can be used to describe the sea surface as a stochastic process, however two conditions (assumptions) must first be met 1) the observation must be stationary and 2) the surface elevations must be Gaussian distributed. As mentioned previously, the random-phase/amplitude model  characterises waves as harmonic and the time-record can be reproduced by the sum of a large number of harmonic waves using a *Fourier analysis*. A Fourier analysis allows values of the amplitude and phase for each frequency to be characterised, which in turn provides the amplitude and phase spectrum for a single time-record. The amplitude spectrum is calculated by averaging many time-records which determines the *average amplitude spectrum*. This approach allows the entire wave record to be characterised but is not statistically relevant and so is not applicable to hydrodynamic modelling. In order for to attain a statistically relevant value, the variance of each wave component must be calculated. The variance is regarded as statistically relevant, as the sum of the variances (i.e. random surface elevation) of the wave components is equal to the variance of the sum of the wave components. In addition, LWT dictates that the energy of waves is proportional to the variance and is therefore an indirect measure of wave energy. This indirect measurement can also be used to determine other important wave components such as wave-induced particle velocity and pressure variations. Important to note that the measurement of variance are discrete and is therefore limited in characterising waves. A continuous scale is more applicable for hydrodynamic modelling purposes and is also closer to representing ocean waves and can be calculated by allowing the frequency interval () to approach zero.

@@

Wave exposure may be modeled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorporates integrative measures of hydrodynamic conditions at a particular site. Cartographic models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurements for larger areas, and has been suggested as a method for predicting macroalgal community structure (???). An example of such a model is the “BioEx model” which was developed by Baardseth and others (1970) to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open), and has been used in mapping marine coastal biodiversity (???). Lindegarth and Gamfeldt (2005) criticised this approach, arguing that the choice of wave exposure method can influence ecological inference. The authors also highlighted the need for objective, reproducible and quantitative studies comparing exposure indices (Lindegarth and Gamfeldt 2005). Other authors, such as Hill et al. (2010), have argued that these simple measures can be improved upon by including bathymetry data which allows the incorporation of diffraction into the calculation. Diffraction is topographically induced variations in wave direction. A model incorporating this complexity was developed by Isæus (2004), and is known as the “simplified wave model” (SWM). The model uses measurements of wind strength, fetch and empirically derived algorithms to mimic diffraction.

Advances in numerical modelling have been founded on physical wave theory which describes how a wave "behaves". This approach is based on a theoretical perspective rather than the need to answer ecological questions. Besides diffraction, numerical models incorporate more complexity by including wind forcing, wave-to-wave interactions and loss of energy due to friction and wave breaking. Numerical models have a variety of applications and are often incorporated within hydrodynamic general circulation models and are used operationally for forecasting the sea state (Group 1988; Booij, Ris, and Holthuijsen 1999; Smith, Sherlock, and Resio 2001). The downside of advanced numerical models is that are computationally intensive which creates limitations for large scale simulations. Therefore, their application along long stretches of variable coastline, inshore environments and ocean-wide simulations are limited due to the poor spatial coverage. However, numerical models can be designed for local or site specific coverage, provided the correct data is available. **...More to be added...**
